Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today

### 2019-11-25 man/ecosystems/ms_ecosystems.Rmd
update infinite version

### Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
* The responses to climate and resilience followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

@@

* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure. Trees were highly sensitive to moisture availability and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term. The response to climate and resilience followed a water-stress gradient. Primary and secondary growth were generally less resilient to drought in xeric than in mesic sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and more humid) high-elevation stands, but absent in the most xeric site.

* Adult trees of the studied species showed high resilience to drought despite being considered relict and at its rear-edge. Yet, the trees exhibited a variable response to the changing climate along a very short climatic gradient and showed that the ecological and geographical rear-edges do not need to meet today in forests with strong land-uses legacies. Forest resilience including that of regeneration needs to be monitored at xeric locations (*i.e.* the low altitudinal rear-edge) under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.

@@

## Introduction
The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range (*i.e.* centre-periphery hypothesis) [@Sagarin2002; @Pironon2016]. It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance and thus higher risk of extinction than those populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginallity does not correspond with ecological marginallity [@VilaCabreraJump2019GreaterGrowth].

Climate change is expected to cause important changes in the distribution and abundance of plant communities, and there are already obvious signs of processes associated with increased intensity and duration of droughts [@Allen2010]. In this respect, several studies have warned on the rapid extinction of plant species in the near future, especially in mountain areas of southern Europe [*e.g.* @Engler2011]. Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].

Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016]. There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a]. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. Yet it is crucial to accurately characterize the rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [@Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services.

Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018]. The resilience to drought of species is better assessed when several approaches like dendrocronology and remote sensing are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual tree-ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such as drought. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018]. Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

We used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd. (*melojo oak*) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades. Our main hypothesis is that these relict populations driven by historical land-use are particularly vulnerable to climate change at their present climatic (either altitudinal or latitudinal) rear-edge, and will show low resilience after extreme drought particularly at the dry edge along climatic gradients at the small scale within the rear-edge. To address this hypothesis our specific aims were: (*i*) To quantify how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in its rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) within the rear edge to detect vulnerability to climate change along climatic gradients at the small spatial scale. To achieve goal (*iii*) and characterize variability in the forest response to drought within the rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.

## Materials and methods 
### Tree species and study site 
*Quercus pyrenaica* Willd. (*melojo* oak) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].

The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability set by altitude. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002]. There are eight melojo oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Today, *Q. pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG]. These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the rear-edge [@GeaIzquierdo2013].

### Drought episodes
The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and to analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below that threshold for at least two consecutive months. For each drought event we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

@@

### Field sampling and dendrochronological methods to assess tree individual resilience
Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled aspect. Sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct for significance.

@@

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010]. Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976]. The statistical quality of each chronology was checked via the expressed population signal. A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

@@

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of existing abundant historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).

@@

### Assessing ecosystem and tree individual resilience to drought
To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011. The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Table S3).

@@

### Analysis of radial growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites and despite competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3). Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid, site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies. Tree-growth reduction followed a moisture-gradient. The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest for the wettest site (Figure 3).

>>>>

### Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
* The growth response to climate and resilience followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

@@

* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure. Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events in the short- and the long-term. The response to climate and resilience followed a water-stress gradient. Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and more humid) high-elevation stands, but absent in the most xeric site.

* Adult oak trees showed high resilience to drought despite being considered relict and at its rear-edge. The trees exhibited a variable response to the changing climate along a very short climatic gradient and showed that the ecological and geographical rear-edges do not need to meet today in forests with strong land-uses legacies. Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios bearing in mind that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.

@@

## Introduction
The response of species to changing environments is likely to be largely determined by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range (*i.e.* centre-periphery hypothesis) [@Sagarin2002; @Pironon2016]. It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance and thus higher risk of extinction than those populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].

Climate change is expected to cause important changes in the distribution and abundance of plant communities, and there are already obvious signs of processes associated with increased intensity and duration of droughts [@Allen2010]. In fact, there has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a]. Under this climatic change context, it is reasonable to expect that population loss and range retractions in boreal, temperate and Mediterranean species should be seen in the lowest latitudes, altitudes and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005]. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of species to predicted climatic changes [@Jump2010]. Yet it is crucial to accurately characterize the ecological rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services. The resilience to drought of species is better assessed when several approaches like dendrocronology and remote sensing are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual tree-ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [*e.g.* @Gazol2018; @PenaGallardo2018]. Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

In this study we used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd. (*melojo oak*) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades. Our main hypothesis is that these relict populations driven by historical land-use are particularly vulnerable to climate change at their present climatic (either altitudinal or latitudinal) rear-edge, and will show low resilience after extreme drought particularly at the dry edge along climatic gradients at the small scale within the rear-edge. To address this hypothesis our specific aims were: (*i*) To quantify how recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in its rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) within the rear edge to detect vulnerability to climate change along climatic gradients at the small spatial scale. To achieve goal (*iii*) and characterize variability in the forest response to drought within the current geographical rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.

```{r, eval=FALSE, echo = FALSE}
# Retales de la intro 
# Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].

# In this respect, several studies have warned on the rapid extinction of plant species in the near future, especially in mountain areas of southern Europe [*e.g.* @Engler2011].

# Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].

# Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016].

# Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [@Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973]. 
```

## Materials and methods 
### Tree species and study site 
*Quercus pyrenaica* forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].

The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations up to 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability set by altitude. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002]. There are eight melojo oak patches (2400 ha) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Today, *Q. pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG]. These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].

### Drought episodes
The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented among the worst in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 and to analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below that threshold for at least two consecutive months. For each drought event we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about their magnitude [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$, 
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

@@

### Field sampling and dendrochronological methods to assess tree individual resilience
Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), which constitute the current low altitudinal limit (CA-Low) and the tree-line (CA-High), respectively in the sampled aspect. Despite the proximity of these two elevations (less than 200 m difference in altitude) the stands greatly differ in their structure and characteristics (Table 1). The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct for significance.

@@

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010]. Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976]. The statistical quality of each chronology was checked via the expressed population signal (EPS). A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

@@

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of existing historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4). We reviewd several documentary sources: historical documents and maps; detailed mining reports; official information about recent wildfires events and forest management practices; livestock farming; traditional irrigation ditches and other studies concerning the socioeconomic dynamics of forest of Sierra Nevada at different scales (see Table S4 for references).

@@

### Assessing ecosystem and tree individual resilience to drought
To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011. The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018]. Resilience metrics for BAI data were additionally computed for the most severe drought events since 1940 (*n* = 8; Table S3) and compared with drought severity.

@@

### Analysis of radial growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites and despite competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3). Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a). Only trees from the southern sites (*i.e.* the more humid aspect) showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid and colder, site (CA-High).

Drought events reduced radial growth for all sites (Figure S2a), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies. Tree-growth reduction followed a moisture-gradient. The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest growth reductions in trees from the wettest site (Figure 3).
